Life in the canopy: Phylogenetic filtering and morphological canalization in epiphytic ferns

### 2018-03-19 manuscript.Rnw
Paste in rest of results and discussion
Still not done going through and changing text results values to Sexpr to grab actual results values.

>>>>

\subsection*{Principal Components Analysis of Trait Data}

Terrestrial and epiphytic species are partly differentiated in trait space, occupying mostly distinct areas but overlapping in the middle regardless of method used (standard or phylogenetic PCA) (\fig \ref{fig:PCA}C, D). The first two PC axes explained \Sexpr{get_result(PCA.results, "Std_PCA_Dim.2", "cumulative_variance", digits=3, signif=TRUE)}\% and \Sexpr{get_result(PCA.results, "Phy_PCA_Dim.2", "cumulative_variance", digits=3, signif=TRUE)}\% of the variation in the trait data when analyzed using standard and phylogenetic PCA, respectively. In both analyses, stipe length and frond width were loaded on PC1, and SLA, degree of frond dissection, and frond length were loaded on PC2 (\fig \ref{fig:PCA}A, B, \tab \ref{supp_table:PCA}). Results for number of pinna pairs differed between the two analyses, tending toward PC2 in the standard PCA and PC1 in the phylogenetic PCA (\tab \ref{supp_table:PCA}).

\subsection*{Phylogenetic Signal}

Most measured traits showed some degree of phylogenetic signal, but the strength varied across traits (\tab \ref{table:phylosig}, \fig \ref{fig:traits_tree}), and for quantitative traits, different results were obtained for $\lambda$ and K. When measured with $\lambda$, most sporophyte traits show phylogenetic signal (as expected under a BM model), with values of $\lambda$ close to 1; only number of pinna pairs had $\lambda$ close to zero (\tab \ref{table:phylosig}a). However, when measured with \K, only frond width and rhizome diameter showed strong phylogenetic signal (more than expected under BM); other traits had values of $\K < 1$ (\tab \ref{table:phylosig}a). All binary gametophyte traits showed significant phylogenetic signal, either similar to (hairs, $\D = \Sexpr{get_result(binary_phylosig.results, "D", "hairs", digits=2, signif=TRUE)}$) or more conserved (gemmmae $\D =  \Sexpr{get_result(binary_phylosig.results, "D", "gemmae", digits=2, signif=TRUE)}$; glands, $\D =  \Sexpr{get_result(binary_phylosig.results, "D", "glands", digits=2, signif=TRUE)}$; morphotype, $\D = \Sexpr{get_result(binary_phylosig.results, "D", "cordate_morph", digits=2, signif=TRUE)}$) than expected under a scenario of evolution by BM (\tab \ref{table:phylosig}b). Phylogenetic generalized linear mixed models (PGLMMs) also indicated significant signal in all gametophyte traits (\tab \ref{supp_table:PGLMM_gameto}). Growth habit showed phylogenetic signal nearly identical to that expected under BM ($\D = \Sexpr{get_result(binary_phylosig.results, "D", "epiphytic", digits=2, signif=TRUE)}$; \tab \ref{table:phylosig}b).

\subsection*{Correlation of Traits with Growth Habit}

Most sporophyte traits, including all those related to plant size, are significantly correlated with growth habit while taking phylogeny into account (\tab \ref{table:PIC_sporo}). Epiphytic species tend to have smaller values for all traits (\tab \ref{table:PIC_sporo}; contrasts were made by subtracting epiphytic from terrestrial values). Short stipe length in particular appears to be strongly correlated with epiphytic growth (phylogenetically independent contrasts, $\pval = \Sexpr{get_result(PIC.results, "pval", "stipe", digits=2, signif=TRUE)}$), as well as frond width ($\pval = \Sexpr{get_result(PIC.results, "pval", "width", digits=2, signif=TRUE)}$) and degree of laminae dissection ($\pval = \Sexpr{get_result(PIC.results, "pval", "dissection", digits=2, signif=TRUE)}$). For the gametophyte traits, only morphotype (non-cordate vs. cordate) was significantly correlated with growth habit (Pagel's test of correlated evolution, $\pval = 0.0011$; \tab \ref{table:pagel_gameto}). Quantitatively similar results were obtained using PGLMMs, except that no significant relationship was observed between rhizome diameter and growth habit, whereas SLA did show a significant relationship with growth habit (\tabs \ref{supp_table:PGLMM_sporo}, \ref{supp_table:PGLMM_gameto}).

\subsection*{Functional and Phylogenetic Diversity}

On a single-trait basis, distribution of community-weighted mean values (CWMs) showed significant correlation with elevation for some traits, but no significant relationship was detected for any traits with minimum RH (\fig \ref{fig:CWM}A). SLA decreased with elevation for epiphytic (linear model, $\rval^2 = 0.6787$, $\pval <0.001$) and terrestrial (linear model, $\rval^2 = 0.8118$, $\pval <0.001$) communities. Degree of lamina dissection decreased with elevation in epiphytes (linear model, $\rval^2 = 0.3634$, $\pval = 0.0104$), but not terrestrial communities. Stipe length showed a very weak, but significant, increase with elevation for epiphytic communities (linear model, $\rval^2 = 0.2721$, $\pval = 0.0317$). Rhizome diameter decreased with elevation for terrestrial communities (linear model, $\rval^2 = 0.238$, $\pval = 0.047$), but increased slightly in epiphytes (linear model, $\rval^2 = 0.3525$, $\pval = 0.0120$).

We observed a total of 104 species across all plots, including 48 epiphytic and 56 terrestrial species. Mean species richness was $11.12 \pm 4.6$ \si{\species} per plot for epiphytes and $12.35 \pm 5.0$ \si{\species} per plot for terrestrial species (not significantly different; one sided \tval-test, $\pval = 0.23$; \fig \ref{fig:FDIV}A). Species richness increased linearly with elevation and minimum RH for epiphytes, but not terrestrial communities (\fig \ref{fig:FDIV}B). Trait richness (FRic) was lower in epiphytic vs. terrestrial communities overall (one sided \tval-test, $\pval < 0.001$), whereas functional evenness (FEve) and diversity (FDiv) did not differ between the two. FEve decreased with elevation, but not minimum RH, in terrestrial communities (\fig \ref{fig:FDIV}A). Epiphytic communities showed a trend of increasing trait richness along the environmental gradient, with greater FRic at high elevations (linear model, $\rval^2 = 0.5836$, $\pval < 0.001$) and plots with greater minimum RH (linear model, $\rval^2 = 0.6626$, $\pval = 0.009$). Epiphytic communities were more phylogenetically clustered than terrestrial communities overall when measured using MNTD (one sided \tval-test, $\pval < 0.001$) but not MPD ($\pval = 0.30$; \fig \ref{fig:FDIV}B). No significant trend was detected in phylogenetic community structure with environment in either terrestrial or epiphytic communities (\fig \ref{fig:FDIV}A).

%%%%%%%%%%%%%%%%%%%%%%%%%%% Discussion %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Discussion}

Here, we show for the first time using a broadly sampled phylogenetic dataset that morphological traits related to epiphytic growth are strongly evolutionarily conserved in ferns. The canalized morphology of epiphytic ferns suggests that this habitat strongly filters for a specialized growth form. We find that communities of epiphytic ferns are more clustered functionally and phylogenetically relative to terrestrial species, again supporting the role of environmental filtering in determining epiphytic fern community structure. Our results are similar between gametophyte and sporophyte traits, suggesting that morphology of each life stage is important for epiphytic growth.

\subsection*{Effect of Phylogeny on Analysis of Epiphytic Traits}

Although several other studies have addressed the relationship of functional traits with growth habit in ferns \citep[\eg, ][]{Hietz1998,Kluge2007,WatkinsJr.2010}, few have accounted for evolutionary history. We find that growth habit shows phylogenetic signal, as do several of the traits associated with it. In one of the few previous studies to analyze phylogenetic signal in functional traits of ferns, \citet{Zhang2014b} found that only one of 16 traits showed significant phylogenetic signal. The only trait in common between the current study and that of \citet{Zhang2014b} is SLA, which did not show phylogenetic signal in either study. Given the high degree of phylogenetic signal we observe in most traits and that at least some of the traits between our two studies are likely to be correlated, it is surprising that \citet{Zhang2014b} did not detect phylogenetic signal in more traits. However, \citet{Zhang2014b} only used \K, which tended to indicate less phylogenetic signal than $\lambda$ in our study. In simulation studies, $\lambda$ has lower error (both Type I and Type II) than \K, which can be inconsistent unless the strength of BM is great \citep{Munkemuller2012}. Our results indicate the importance of including a phylogenetic component, and including multiple measures of phylogenetic signal, when analyzing traits related to epiphytic growth in ferns.

\subsection*{Sporophyte Traits}

The sporophyte traits that differed most strongly between epiphytic and terrestrial taxa were related to overall size, including stipe and frond length, frond width, and rhizome diameter. Stipe length in particular differed strongly between epiphytes and terrestrial species (\fig\ref{fig:CWM}B, \tab\ref{table:PIC_sporo}). Stipe length may be a critical trait controlling whole leaf hydraulic conductance; since ferns have resistive stipes in general, species growing in water-limited habitats are expected to decrease stipe length to minimize resistance \citep{WatkinsJr.2010}. Our results support previous non-phylogenetically controlled studies that found a prevalence of short stipes in epiphytic ferns \citep{WatkinsJr.2010,Creese2011}. However, given that many of these size-related traits scale together in ferns \citep{Arcand2008,Creese2011}, careful experimental manipulations are needed to distinguish the role (or lack thereof) of each in epiphytic growth.

Surprisingly, SLA was only slightly lower in epiphytes relative to terrestrial species (\tab\ref{table:PIC_sporo}, \fig\ref{fig:CWM}B). SLA is an important part of the leaf economic spectrum, which posits that leaf traits including SLA, N and C content, and leaf life span vary along well-defined axes such that plants generally fall into one of two categories: fast-growing species with high SLA, low nutrient content, and short life spans vs. slow-growing species with low SLA, high nutrient content, and long life spans \citep{Wright2004a}. We hypothesized that epiphytes would have lower SLA, given their more extreme growing conditions. Indeed, the species with the lowest SLA values were epiphytes (\eg, \textit{Selliguea plantaginea} Brack., \SI{8.7}{\square\meter\per\kilo\gram}; \textit{Davallia anderssonii} Mett. ex Kuhn, \SI{8.43}{\square\meter\per\kilo\gram}), but this relationship was not significant across the dataset (\tab\ref{table:PIC_sporo}, \fig\ref{fig:CWM}). The tradeoffs posited in the leaf economics spectrum have been most thoroughly documented in global studies including the widest possible range of taxa and environments, and do not always hold up in local studies \citep{Petter2016}. For example, \citet{Zhu2016} did not find a significant relationship between SLA and leaf life span in a survey of 16 ferns across a disturbance (light) gradient in China. It is possible that other small-scale effects (\eg, microclimate) obscure this pattern in local studies such as ours and that of \citet{Zhu2016}. This is supported by the strong, parallel relationship in CWM values of SLA with elevation in both epiphytes and terrestrial species (\fig\ref{fig:CWM}A).

\subsection*{Gametophyte Traits}

Only gametophyte morphotype was found to have a significant association with epiphytic growth. Morphotype was previously observed to correlate with life history and habitat in ferns: terrestrial species tend to have short-lived ($< \SI{1}{\year}$), cordate gametophytes that establish following disturbance and rapidly produce sporophytes, whereas epiphytic species tend to have non-cordate (\eg, elongate or filamentous) gametophytes that persist over multiple growing seasons and produce sporophytes more slowly \citep{WatkinsJr.2007a,Farrar2008}. The complex, three-dimensional structure of non-cordate gametophytes slows their drying rates, and may be an adaptation to drought-prone epiphytic habitats \citep{WatkinsJr.2007,Pittermann2013}, and promote out-crossing \citep{Farrar2008}. Our study is the first to our knowledge to demonstrate that the correlation between non-cordate morphology and epiphytism in ferns is significant while controlling for phylogeny. \cite{Sundue2015} found that non-cordate morphology is significantly correlated with increased diversification rates within Polypodiaceae, consistent with the hypothesis that this trait promoted diversification within the epiphytic niche.

We did not find that gemmae production was linked with epiphytism. Gemmae are asexual propagules that allow gametophytes to persist over multiple growing seasons and attain large population sizes \citep{Farrar2008}. Gemmae production should be a useful trait in an epiphytic context because it may allow gametophytes to increase their surface area and hedge against population loss due to extreme drying events \citep{Farrar2008}. Gemmae can also function sexually by producing antherida in the presence of antheridiogen secreted by other gametophytes \citep{Emigh1977}, which may confer a selective advantage in the epiphytic environment by enabling outcrossing \citep{Dassler2001}. It is possible that some, but not all, epiphytic ferns rely on such strategies, and that they are not under similarly strong selection as other traits such as stipe length and gametophyte morphotype. Interestingly, when we classified `widespread' species as those with gametophytes occurring more than \SI{200}{\meter} above or below the minimum or maximum elevational range of conspecific sporophytes (data not shown), we found that gemmae production is correlated with widespread growth (log likelihood dependent model = \num{-54.5}, log likelihood independent model = \num{-63.6}, $\pval = 0.001$, Pagel's test of correlated evolution). Thus, it seems that gemmae do function in long-term gametophyte persistence, but may not be required for epiphytic growth.

Presence of hairs and glands has strong phylogenetic signal (\tabs\ref{table:phylosig}, \ref{supp_table:PGLMM_gameto}), and these traits are correlated with certain taxonomic groups \citep{Stokey1951,Nayar1971}. Although hairs have been hypothesized to reduce rates of drying by increasing the boundary layer and holding external water \citep{WatkinsJr.2007}, we did not detect a correlation between hair production and epiphytic growth. Similarly, production of glands does not appear to be correlated with epiphytic growth. Wax-secreting glands are prevalent in both sporophytes and gametophytes of many notholaenid ferns, a xeric-adapted clade, but their functional significance has not been established \citep{Johnson2012}. Further tests are needed to determine if such glands serve a functional role in preventing water loss in desert ferns and epiphytes.

\subsection*{Environmental Filtering}

We observed the predicted trend of decreased trait richness (FRic) in epiphytic relative to terrestrial communities overall, and a significant positive correlation of FRic with humidity in epiphytes (\ie, decreased FRic at drier, low elevation plots and increased FRic at more humid, high elevation plots; \fig\ref{fig:FDIV}A). However, we did not observe similar patterns in the other two measures of functional diversity, evenness (FEve) and divergence (FDiv) (\fig\ref{fig:FDIV}B). Indeed, FEve decreased with elevation (but not minimum RH) in terrestrial communities (\fig\ref{fig:FDIV}A). Overall, the lack of a clear pattern in functional evenness and divergence suggests that within each trait pool, there is no significant difference in trait spacing between epiphytes and terrestrial species. This could be maintained by niche partitioning within epiphytic and terrestrial habitats. Epiphytic habitats are highly heterogeneous, and microclimatic conditions vary with height along the host and distance from the trunk \citep{Hietz1995}. The vertical stratification of epiphytic species into different zones along height gradients in host trees is well documented \citep{Zotz2007}. In a study comparing terrestrial and epiphytic ferns along an elevational gradient from \SIrange{30}{3000}{\meter} in Costa Rica, \citet{WatkinsJr.2006a} found that epiphytes tended to sort into low-trunk and canopy zones, with only 18\% found in both, and no species that overlapped between canopy and terrestrial habitats. It is possible that epiphytic growth per se limits overall morphological variation in epiphytic ferns, but that species with different traits are then sorted into different niches.

In addition to their decreased functional diversity, we also found that epiphytic communities are more phylogenetically clustered relative to terrestrial communities overall (\fig\ref{fig:FDIV}B). However, we did not detect significant trends in phylogenetic community structure of epiphytic and terrestrial communities along the elevational gradient (\fig\ref{fig:FDIV}A). It is possible that the scale of the current study (from \SIrange{200}{1200}{\meter}) prevented the detection of such trends. When the plots in this study were analyzed together with additional plots from Tahiti reaching a maximum altitude of just over \SI{2000}{\meter}, \citet{Nitta2017}  detected increasing phylogenetic clustering in fern sporophyte communities (epiphytes and terrestrial species together) with elevation. \cite{Kluge2011} also compared phylogenetic and trait diversity between epiphytic and terrestrial ferns over an elevational gradient from \SIrange{100}{3400}{\meter} in Costa Rica, and found trait overdispersion at mid-elevations for epiphytes but no trends in phylogenetic diversity for either group. They accounted this to evolutionary plasticity in traits, but did not quantify the degree of phylogenetic signal in the traits they studied.

\subsection*{Conclusion and Future Directions}

We have demonstrated that growth habit itself, and several traits associated with it, show significant phylogenetic signal in ferns. Furthermore, we find that epiphytes have restricted morphologies both on a per-species basis and at the community level, and that epiphytic communities are phylogenetically clustered relative to terrestrial ones. Finally, we observed drier conditions in the canopy relative to terrestrial habitats. Taken together, our results support a scenario of environmental filtering acting on conserved traits to limit functional and phylogenetic diversity in epiphytic ferns. The generality of this pattern could be further tested by investigating the functional traits of other groups with large epiphytic radiations such as orchids and bromeliads.

Recent studies \citep[\eg,][]{Sundue2015,Testo2016a,Lehtonen2017} have called into question the scenario of rapid Cretaceous radiations of epiphytic ferns posited by \cite{Schuettpelz2009} by failing to detect increased rates of diversification associated with epiphytic growth. Epiphytism itself is likely a complex trait requiring the evolution of multiple associated traits, and the timing and order of the evolution of such traits may be related to the complex diversification dynamics observed in these studies. Our study identifies several conserved morphological traits (\eg, stipe length, frond dissection, gametophyte morphotype) that may belong to this suite of epiphytic traits. We hope future studies will build on our results by analyzing these traits across much more densely sampled phylogenies \citep[\eg][]{Testo2016a} to dissect how each may be related to diversification and growth habit. Additionally, future studies should focus on determining in more detail how niche space is partitioned within epiphytic vs. terrestrial habitats, which is clearly a major dividing line for ferns.
